GPU-friendly Stroke Expansion

### 2024-04-29 paper.tex
add missing units
Fix citations
Changed the format to "authoryear". The text now uses \citet for
author[year] type in-sentence citations and \citenum for parenthetical
citations. This conforms to the ACM style rules described on
https://www.acm.org/publications/authors/reference-formatting.

\documentclass[sigconf, authordraft]{acmart}
\usepackage[linesnumbered]{algorithm2e}
\usepackage[format=plain, labelfont={bf}, textfont=normal]{caption}
\usepackage{booktabs,xcolor}
\begin{document}
\title{GPU-friendly Stroke Expansion}
\author{Raph Levien}
\author{Arman Uguray}
\affiliation{
    \institution{Google}
    \city{San Francisco}
    \state{CA}
    \country{USA}
}

@@

Stroke expansion is the process of generating an outline that, when filled, produces the \emph{stroke} of a given input path. Informally, a line of the same width as the stroke is swept along the path, normal to it, and also decorated with joins and caps. Though stroke rendering is referenced in many 2D graphics standards, it has lacked a precise formal definition until recent work, primarily \cite{Nehab2020} and \cite{Kilgard2020}.

The \cite{Nehab2020} paper gives a comprehensive survey of techniques for stroke expansion, and an algorithm which, for a number of reasons, is only suitable for implementation on CPU (though adapting it to GPU is listed as promising future work). It classifies techniques into \emph{local,} where each path segment generates closed geometry (which may be triangles or other primitives), and \emph{global,} where the overall result is a closed outline of the stroked path, but the partial result from each segment is in general open. Our technique is considered global in this scheme, yet allows independent processing of each path segment.

@@

We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins. Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins. The two notions are equivalent for sufficiently well-behaved input, in particular when the curvature measured at endpoints of path segments does not exceed the reciprocal of the half linewidth. As described in detail in \cite{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints. Standards for graphics formats, such as SVG\cite{Svg2}, also provide an ``out,'' enabling the weaker behavior. While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.

A great number of rendering techniques for strokes have been proposed. Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke. Included in this category is polar stroking \cite{Kilgard2020}, which proposes a GPU-friendly subdivision scheme based on an angle step error. Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering. Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \cite{Nehab2020}. In both cases, the outline contains approximations of parallel curves of the input curve segments, and, as proposed in \cite{Nehab2020}, their evolutes when targeting strong correctness. For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \cite{Nehab2020}.

The subproblem of approximating parallel curves has also spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU. An example of a reasonably good algorithm that computes flattened parallel curves is \cite{Yzerman2020}, as used in the Blend2D rendering library. For producing curved outlines, the Tiller and Hanson algorithm\cite{Tiller1984} is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers. The quadratic Bézier version is adequate, and forms the basis of the algorithm in \cite{Nehab2020}. A variant is also used in Skia\cite{Skia}; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.

@@

There are generally three approaches to such curve approximation. The most straightforward but also least efficient is ``cut then measure,'' usually combined with adaptive subdivision. In this technique, a candidate approximate curve is produced, then the error is measured against the source curve, usually by sampling a number of points along both curves and determining a maximum error (or perhaps some other error norm). If the error is within tolerance, the approximation is accepted. Otherwise, the curve is subdivided (usually at $t = 0.5$) and each half is recursively approximated. A substantial fraction of all curve approximation methods in the literature are of this form, including \cite{Nehab2020}. The main disadvantage is the cost of computing the error metric. Another risk is underestimating the error due to inadequate sampling; this is a particular problem when the source curve contains a cusp.

@@

By far the most efficient approach is an \emph{invertible} error metric. In this approach, the error metric has an analytic inverse, or at least a good numerical approximation. Because the metric is invertible, it can predict the number of subdivisions needed, as well as the parameter value for each subdivision. If the error metric is accurate, then approximation is near-optimal. One example of an invertible error metric is angle step, used in polar stroking\cite{Kilgard2020}; the number of subdivisions is the total angle subsumed by the curve divided by the angle step size, and the parameter value for each subdivision is the result of solving for a tangent direction. Another widely used invertible error metric is Wang's formula (\cite{Goldman2003}, Section 5.6.3), which gives a bound on the flattening error based on the second derivative of the curve. This metric is conservative but works well in practice; among other applications, it is used in Skia for path flattening. One limitation of Wang's method is that it only applies to the flattening of filled outlines; when applied naively to generation of parallel curves, it can undershoot substantially, especially near cusps.

@@

It is common to approximate cubic Béziers to some intermediate curve format more conducive to offsetting and flattening. A number of published solutions (\cite{Yzerman2020}, \cite{Nehab2020}) use quadratic Béziers, as it is well suited for computation of parallel curves. Even so, this curve has some disadvantages. In particular, it cannot model an inflection point, so the source curve must be subdivided at inflection points.

@@

The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by Wieleitner well over a hundred years ago\cite{Wieleitner1907}, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length. TODO: present closed-form solution.

@@

Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering. Another choice would be uniform steps in tangent angle, as chosen by polar stroking\cite{Kilgard2020}. However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering. It should be noted, to avoid the undershoot at low curvature, both the Skia\cite{Skia} and Rive\cite{Rive} renderers use a hybrid of the Wang and polar stroking error metrics.

@@

For a convex curve, the optimal flattening is tractable to compute, though not necessarily very fast. When a curve is monotonic, error is also monotonic with respect to both subdivision points. Thus, any standard root finding technique can find a subdivision with a given error (bisection is the best known, but the ITP method \cite{Oliveira2020} is better). An similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments. Subdividing at the inflection point adds at most one additional point; it is possible to optimize further, but we do not consider that (though the near-optimal flattening we will later present is not subject to this constraint).

@@

The problem of approximating a curve by a sequence of arc segments has extensive literature, but none of the published solutions are quite suitable for our application. The specific problem of approximating an Euler spiral by arcs is considered in \cite{Meek2004} using a ``cut then measure'' adaptive subdivision scheme, but their solution has poor quality; it scales as $O(1/n^2)$, while $O(1/n^3)$ is attainable. The result was improved ``slightly'' by \cite{Narayan2014}. The literature also contains optimal results, namely \cite{Maier2014} and \cite{Nuntawisuttiwong2021}, but at considerable cost; both approaches claim $O(n^2)$ time complexity. The through-line for all these results is that they are solving a harder problem: adopting the constraint that the generated arc sequence is $G^1$ continuous. While desirable for many applications, this constraint is not needed for rendering a stroke outline. Even with this constraint relaxed, the angle discontinuities of an arc approximation are tiny compared to flattening to lines.

@@

In the principled, correct specification for stroking\cite{Nehab2020}, parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width. When it does, additional segments must be drawn, including evolutes of the original curve. In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques. By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve\cite{Yoshida2012}.

@@

Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type, more tractable for evaluating parallel curve. Computing parallel curves directly on cubic Bézier curve segments is not very tractable. In particular, the widely cited Tiller-Hanson algorithm\cite{Tiller1984} performs well for quadratic Béziers but significantly worse for cubics.

@@

A typical pattern for converting from one curve type to another is \emph{adaptive subdivision.} (TODO: is there a good general reference for this?) An approximate curve is found in the parameter space of the target curve family. The error of the approximation is measured. If the error exceeds the specified tolerance, the curve is subdivided (typically at $t = 0.5$), otherwise the approximation is accepted. Subdivisions are also indicated at special points; for example, since quadratic Béziers cannot represent inflection points, and geometric Hermite interpolation is numerically unstable if the input curve is not convex, lowering to quadratic Béziers also requires calculation of inflection points and subdividing there. A good example of this pattern is Nehab\cite{Nehab2020}. One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to compute those for additional subdivision.

@@

A key step in approximating one curve with another is evaluating the error of the approximation. A common approach (used in \cite{Nehab2020} among others) is to generate the approximate curve, the measure the distance, often by sampling at multiple points on the curve. All this is potentially slow, with the additional risk of underestimating the error due to undersampling.

@@

Given tangent angles relative to the chord, finding the Euler spiral segment that minimizes total curvature variation is a form of geometric Hermite interpolation. There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent\cite{Kimia2003}, bisection\cite{Walton2009}, or Newton iteration\cite{Connor2014}. A more direct approach is to approximate the function as a reasonably low-order polynomial in terms of the endpoint angles. Our approach is to use a 7th order polynomial, which is more precise than the 3rd order polynomial proposed in \cite{Reif2021}, at only modest increased cost. The exact coefficients used are presented in Appendix \ref{appendix:gh}.

@@

As recommended in both \cite{Nehab2020} and \cite{Kilgard2020}, the rendering of a cusp with infinite curvature matches that of a near-cusp. The code to detect near-zero derivatives and re-evaluate is a small amount of non-divergent logic, unlike the additional ``regularization'' pass proposed in Section 3.1 of \cite{Nehab2020} to handle special cases.

@@

Even when the input curve is smooth, its parallel curve contains a cusp when the (signed) curvature equals the offset distance (the half-width of the stroke). In published techniques, dealing with these cusps is a nontrivial effort, and involves numerical methods that are not GPU friendly. In particular, detecting locations in the cubic Bézier where the curvature crosses a given quantity is a medium-degree polynomial, and in general requires numerical techniques for root finding. In the approach of \cite{Nehab2020}, the root finder not only requires a hybrid Newton/bisection method, which requires iteration, but is also recursive in that it uses the roots of a polynomial of one degree lower as a subroutine. In general, polynomials up to cubic can be considered GPU-friendly, while finding cusps in cubic Béziers requires a higher degree.

@@

We present two versions of our stroking method: a sequential CPU stroker and the GPU compute shader outlined in Section \ref{section:gpu-impl}. Both versions can generate stroked outlines with line and arc primitives. We focused our evaluation on the number of output primitives and execution time, using the \emph{timings} dataset provided by Nehab2020\cite{Nehab2020}, in addition to our own curve-intensive scenes to stress the GPU implementation. We present our findings in the remainder of this section.

@@

We measured the timing of our CPU implementation, generating both line and arc primitives, against Skia's stroker, which outputs a mix of straight lines and quadratic Bézier primitives. The results are shown in Figure~\ref{fig:timings}. The time shown is the total for test files from the \emph{timings} dataset of \cite{Nehab2020}, and the tolerance in all cases was 0.25. TODO: measure against other implementations.

Similarly to \cite{Nehab2020}'s observations, Skia is the fastest stroke expansion algorithm we measured. The speed is attained with some compromises, in particular an imprecise error estimation that can sometimes generate outlines exceeding the given tolerance. We confirmed that this behavior is still present. The Nehab paper proposes a more careful error metric, but a significant fraction of the total computation time is dedicated to evaluating it.

@@

We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, and two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}. We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}\cite{wgpu} framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).

We instrumented our pipeline to obtain fine-grained pipeline stage execution times using the GPU \emph{timestamp query} feature supported by both Metal and Vulkan. This allowed us to collect measurements for the curve flattening and input processing stages in isolation. We ran the compute pipeline on the \emph{timings} SVG files provided by Nehab\cite{Nehab2020}. The largest of these SVGs is \emph{waves} (see rendering \emph{(a)} in Figure \ref{fig:renderings}) which contains 42 stroked paths and a total of 13,308 path segments.

We authored two additional test scenes in order stress the GPU with a higher workload. The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus. We used two variants of this scene with \emph{butt} and \emph{round} cap styles. For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark\cite{MotionMark}, which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths. We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively. Table~\ref{table:bump-counts} shows the payload sizes of our largest test scenes.

@@

\begin{figure}
    \includegraphics[scale=0.69]{nehab_gpu_timings}
    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the Nehab\cite{Nehab2020} \emph{timings} data set. The timings are shown in milliseconds. The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs. Mali-G78 (the lowest end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in ~3ms. The compute shader executes at least an order of magnitude faster on the other GPUs.}
    \Description{A stacked bar chart of the GPU timings for stroke expansion from Nehab\cite{Nehab2020} timings SVGs}
    \label{fig:nehab-gpu-timings}
\end{figure}

@@

Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets. The entire Nehab\cite{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit. Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms. We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested. We also confirmed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.

@@

Our GPU implementation targets graphics APIs that do not support dynamic memory allocations in shader code. This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.

We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment. We found that Wang's formula works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}). For parallel curves, Wang's theorem does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth. We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.

@@

\begin{table}
    \caption{The estimated and actual \emph{line soup} buffer sizes for scenes rendered at 2088x1600 resolution. Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth. In the worst case, our buffer size estimator overshoots by a factor ~4.}
    \begin{tabular}{|l|c|c|}
    \hline
        \multicolumn{1}{|c|}{\textbf{Test}} & Actual Size & Estimated Size \\
    \hline
        roads.svg  & 958.78 KB & 2.45 MB  \\
        waves.svg  & 9.37 MB   & 10.44 MB \\
        mmark-70k  & 23.81 MB  & 100.79   \\
        mmark-120k & 40.86 MB  & 173.56   \\
    \hline
    \end{tabular}
    \label{table:bump-estimate-mem}
\end{table}

@@

\begin{itemize}
    \item The GPU implementation sequentializes some work that could be parallel, and is thus not as well load-balanced as it might be. An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph\cite{Patel2024}. This structure is appealing because load balancing is done by hardware.

>>>>

\documentclass[sigconf, authordraft]{acmart}
\usepackage[linesnumbered]{algorithm2e}
\usepackage[format=plain, labelfont={bf}, textfont=normal]{caption}
\usepackage{booktabs,xcolor}
\usepackage{natbib}
\setcitestyle{authoryear}
\begin{document}
\title{GPU-friendly Stroke Expansion}
\author{Raph Levien}
\author{Arman Uguray}
\affiliation{
    \institution{Google}
    \city{San Francisco}
    \state{CA}
    \country{USA}
}

@@

Stroke expansion is the process of generating an outline that, when filled, produces the \emph{stroke} of a given input path. Informally, a line of the same width as the stroke is swept along the path, normal to it, and also decorated with joins and caps. Though stroke rendering is referenced in many 2D graphics standards, it has lacked a precise formal definition until recent work, primarily \citet{Nehab2020} and \citet{Kilgard2020}.

The \citet{Nehab2020} paper gives a comprehensive survey of techniques for stroke expansion, and an algorithm which, for a number of reasons, is only suitable for implementation on CPU (though adapting it to GPU is listed as promising future work). It classifies techniques into \emph{local,} where each path segment generates closed geometry (which may be triangles or other primitives), and \emph{global,} where the overall result is a closed outline of the stroked path, but the partial result from each segment is in general open. Our technique is considered global in this scheme, yet allows independent processing of each path segment.

@@

We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins. Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins. The two notions are equivalent for sufficiently well-behaved input, in particular when the curvature measured at endpoints of path segments does not exceed the reciprocal of the half linewidth. As described in detail in \citet{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints. Standards for graphics formats, such as SVG[\citenum{Svg2}], also provide an ``out,'' enabling the weaker behavior. While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.

A great number of rendering techniques for strokes have been proposed. Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke. Included in this category is polar stroking[\citenum{Kilgard2020}], which proposes a GPU-friendly subdivision scheme based on an angle step error. Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering. Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \citet{Nehab2020}. In both cases, the outline contains approximations of parallel curves of the input curve segments, and, as proposed in \citet{Nehab2020}, their evolutes when targeting strong correctness. For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \citet{Nehab2020}.

The subproblem of approximating parallel curves has also spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU. An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library. For producing curved outlines, the Tiller and Hanson[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers. The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}. A variant is also used in Skia[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.

@@

There are generally three approaches to such curve approximation. The most straightforward but also least efficient is ``cut then measure,'' usually combined with adaptive subdivision. In this technique, a candidate approximate curve is produced, then the error is measured against the source curve, usually by sampling a number of points along both curves and determining a maximum error (or perhaps some other error norm). If the error is within tolerance, the approximation is accepted. Otherwise, the curve is subdivided (usually at $t = 0.5$) and each half is recursively approximated. A substantial fraction of all curve approximation methods in the literature are of this form, including \citet{Nehab2020}. The main disadvantage is the cost of computing the error metric. Another risk is underestimating the error due to inadequate sampling; this is a particular problem when the source curve contains a cusp.

@@

By far the most efficient approach is an \emph{invertible} error metric. In this approach, the error metric has an analytic inverse, or at least a good numerical approximation. Because the metric is invertible, it can predict the number of subdivisions needed, as well as the parameter value for each subdivision. If the error metric is accurate, then approximation is near-optimal. One example of an invertible error metric is angle step, used in polar stroking (\citet{Kilgard2020}); the number of subdivisions is the total angle subsumed by the curve divided by the angle step size, and the parameter value for each subdivision is the result of solving for a tangent direction. Another widely used invertible error metric is Wang's formula (\citet{Goldman2003}, Section 5.6.3), which gives a bound on the flattening error based on the second derivative of the curve. This metric is conservative but works well in practice; among other applications, it is used in Skia for path flattening. One limitation of Wang's method is that it only applies to the flattening of filled outlines; when applied naively to generation of parallel curves, it can undershoot substantially, especially near cusps.

@@

It is common to approximate cubic Béziers to some intermediate curve format more conducive to offsetting and flattening. A number of published solutions (\citet{Yzerman2020}, \citet{Nehab2020}) use quadratic Béziers, as it is well suited for computation of parallel curves. Even so, this curve has some disadvantages. In particular, it cannot model an inflection point, so the source curve must be subdivided at inflection points.

@@

The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago[\citenum{Wieleitner1907}], and has a straightforward closed-form Cesàro representation, curvature as a function of arc length. TODO: present closed-form solution.

@@

Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering. Another choice would be uniform steps in tangent angle, as chosen by polar stroking[\citenum{Kilgard2020}]. However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering. It should be noted, to avoid the undershoot at low curvature, both the Skia[\citenum{Skia}] and Rive[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.

@@

For a convex curve, the optimal flattening is tractable to compute, though not necessarily very fast. When a curve is monotonic, error is also monotonic with respect to both subdivision points. Thus, any standard root finding technique can find a subdivision with a given error (bisection is the best known, but the ITP method \citet{Oliveira2020} is better). An similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments. Subdividing at the inflection point adds at most one additional point; it is possible to optimize further, but we do not consider that (though the near-optimal flattening we will later present is not subject to this constraint).

@@

The problem of approximating a curve by a sequence of arc segments has extensive literature, but none of the published solutions are quite suitable for our application. The specific problem of approximating an Euler spiral by arcs is considered in \citet{Meek2004} using a ``cut then measure'' adaptive subdivision scheme, but their solution has poor quality; it scales as $O(1/n^2)$, while $O(1/n^3)$ is attainable. The result was improved ``slightly'' by \citet{Narayan2014}. The literature also contains optimal results, namely \citet{Maier2014} and \citet{Nuntawisuttiwong2021}, but at considerable cost; both approaches claim $O(n^2)$ time complexity. The through-line for all these results is that they are solving a harder problem: adopting the constraint that the generated arc sequence is $G^1$ continuous. While desirable for many applications, this constraint is not needed for rendering a stroke outline. Even with this constraint relaxed, the angle discontinuities of an arc approximation are tiny compared to flattening to lines.

@@

In the principled, correct specification for stroking[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width. When it does, additional segments must be drawn, including evolutes of the original curve. In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques. By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve[\citenum{Yoshida2012}].

@@

Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type, more tractable for evaluating parallel curve. Computing parallel curves directly on cubic Bézier curve segments is not very tractable. In particular, the widely cited Tiller-Hanson algorithm[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.

@@

A typical pattern for converting from one curve type to another is \emph{adaptive subdivision.} (TODO: is there a good general reference for this?) An approximate curve is found in the parameter space of the target curve family. The error of the approximation is measured. If the error exceeds the specified tolerance, the curve is subdivided (typically at $t = 0.5$), otherwise the approximation is accepted. Subdivisions are also indicated at special points; for example, since quadratic Béziers cannot represent inflection points, and geometric Hermite interpolation is numerically unstable if the input curve is not convex, lowering to quadratic Béziers also requires calculation of inflection points and subdividing there. A good example of this pattern is \citet{Nehab2020}. One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to compute those for additional subdivision.

@@

A key step in approximating one curve with another is evaluating the error of the approximation. A common approach (used in \citet{Nehab2020} among others) is to generate the approximate curve, the measure the distance, often by sampling at multiple points on the curve. All this is potentially slow, with the additional risk of underestimating the error due to undersampling.

@@

Given tangent angles relative to the chord, finding the Euler spiral segment that minimizes total curvature variation is a form of geometric Hermite interpolation. There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent[\citenum{Kimia2003}], bisection[\citenum{Walton2009}], or Newton iteration[\citenum{Connor2014}]. A more direct approach is to approximate the function as a reasonably low-order polynomial in terms of the endpoint angles. Our approach is to use a 7th order polynomial, which is more precise than the 3rd order polynomial proposed in \citet{Reif2021}, at only modest increased cost. The exact coefficients used are presented in Appendix \ref{appendix:gh}.

@@

As recommended in both \citet{Nehab2020} and \citet{Kilgard2020}, the rendering of a cusp with infinite curvature matches that of a near-cusp. The code to detect near-zero derivatives and re-evaluate is a small amount of non-divergent logic, unlike the additional ``regularization'' pass proposed in Section 3.1 of \citet{Nehab2020} to handle special cases.

@@

Even when the input curve is smooth, its parallel curve contains a cusp when the (signed) curvature equals the offset distance (the half-width of the stroke). In published techniques, dealing with these cusps is a nontrivial effort, and involves numerical methods that are not GPU friendly. In particular, detecting locations in the cubic Bézier where the curvature crosses a given quantity is a medium-degree polynomial, and in general requires numerical techniques for root finding. In the approach of \citet{Nehab2020}, the root finder not only requires a hybrid Newton/bisection method, which requires iteration, but is also recursive in that it uses the roots of a polynomial of one degree lower as a subroutine. In general, polynomials up to cubic can be considered GPU-friendly, while finding cusps in cubic Béziers requires a higher degree.

@@

We present two versions of our stroking method: a sequential CPU stroker and the GPU compute shader outlined in Section \ref{section:gpu-impl}. Both versions can generate stroked outlines with line and arc primitives. We focused our evaluation on the number of output primitives and execution time, using the \emph{timings} dataset provided by \citet{Nehab2020}, in addition to our own curve-intensive scenes to stress the GPU implementation. We present our findings in the remainder of this section.

@@

We measured the timing of our CPU implementation, generating both line and arc primitives, against Skia's stroker, which outputs a mix of straight lines and quadratic Bézier primitives. The results are shown in Figure~\ref{fig:timings}. The time shown is the total for test files from the \emph{timings} dataset of \citet{Nehab2020}, and the tolerance in all cases was 0.25. TODO: measure against other implementations.

Similarly to \citet{Nehab2020}'s observations, Skia is the fastest stroke expansion algorithm we measured. The speed is attained with some compromises, in particular an imprecise error estimation that can sometimes generate outlines exceeding the given tolerance. We confirmed that this behavior is still present. The Nehab paper proposes a more careful error metric, but a significant fraction of the total computation time is dedicated to evaluating it.

@@

We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, and two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}. We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}[\citenum{wgpu}] framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).

We instrumented our pipeline to obtain fine-grained pipeline stage execution times using the GPU \emph{timestamp query} feature supported by both Metal and Vulkan. This allowed us to collect measurements for the curve flattening and input processing stages in isolation. We ran the compute pipeline on the \emph{timings} SVG files provided by \citet{Nehab2020}. The largest of these SVGs is \emph{waves} (see rendering \emph{(a)} in Figure \ref{fig:renderings}) which contains 42 stroked paths and a total of 13,308 path segments.

We authored two additional test scenes in order stress the GPU with a higher workload. The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus. We used two variants of this scene with \emph{butt} and \emph{round} cap styles. For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths. We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively. Table~\ref{table:bump-counts} shows the payload sizes of our largest test scenes.

@@

\begin{figure}
    \includegraphics[scale=0.69]{nehab_gpu_timings}
    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the \citet{Nehab2020} \emph{timings} data set. The timings are shown in milliseconds. The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs. Mali-G78 (the lowest end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in ~3ms. The compute shader executes at least an order of magnitude faster on the other GPUs.}
    \Description{A stacked bar chart of the GPU timings for stroke expansion from \citet{Nehab2020} timings SVGs}
    \label{fig:nehab-gpu-timings}
\end{figure}

@@

Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets. The entire \citet{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit. Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms. We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested. We also confirmed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.

@@

Our GPU implementation targets graphics APIs that do not support dynamic memory allocation in shader code (in contrast to CUDA, which provides the equivalent of \emph{malloc}). This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.

We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment. We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}). For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth. We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.

@@

\begin{table}
    \caption{The estimated and actual \emph{line soup} buffer sizes for scenes rendered at 2088x1600 resolution. Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth. In the worst case, our buffer size estimator overshoots by a factor ~4.}
    \begin{tabular}{|l|c|c|}
    \hline
        \multicolumn{1}{|c|}{\textbf{Test}} & Actual Size & Estimated Size \\
    \hline
        roads.svg  & 958.78 KB & 2.45 MB    \\
        waves.svg  & 9.37 MB   & 10.44 MB   \\
        mmark-70k  & 23.81 MB  & 100.79 MB  \\
        mmark-120k & 40.86 MB  & 173.56 MB  \\
    \hline
    \end{tabular}
    \label{table:bump-estimate-mem}
\end{table}

\section{Conclusion}

TODO:

@@

\begin{itemize}
    \item The GPU implementation sequentializes some work that could be parallel, and is thus not as well load-balanced as it might be. An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph[\citenum{Patel2024}]. This structure is appealing because load balancing is done by hardware.

    \item The pre-allocation requirement for bump-allocated line soup buffer is a limitation due to today's graphics APIs. A more accurate and optimized buffer size estimation heuristic is considered future work. It may also be interesting to explore whether graphics APIs can be extended to facilitate GPU-driven scheduling of the compute workload under a bounded memory footprint.
